---
title: "Rewriting R Code"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rewriting R Code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes transformations to R code that preserve the original
semantics but make analysis easier by standardizing the syntax for certain
tasks.

## Inserting `return()` Statements

<!--
By default, R functions return the value of the last expression in their body.
Since the number `42` is the last expression in the function

```{r, eval = FALSE}
meaning_of_life = function(x, y, z) {

  # ...

  42
}
```

the function returns `42`.

The default behavior can be overridden with a `return()` statement, which
returns its argument and exits the enclosing function immediately. This is
useful for exiting a function early if some condition is met. For instance, the
function

```{r, eval = FALSE}
square_negatives = function(x) {
  if (x < 0)
    return (x^2)

  x
}
```

returns `x^2` if `x` is negative; otherwise the function returns `x`.

Although it's syntactically valid to write a `return()` statement as the last
expression in a function, this is discouraged in the R community because it's
less concise than relying on the implicit behavior.
-->

Functions return the value of their last evaluated expression. An explicit
`return()` is only required to exit a function early. Using `return()` at the
end of a function is discouraged within the R community because it makes code
less concise.

Identifying return values is necessary for code generation and also for data
flow analyses. We can rewrite code so that all return values have an explicit
`return()`. Then it's easy to find return values by class.

When the last expression in a function is a literal, symbol, or call, the
expression can just be wrapped with a `return()`. Rewriting is not so
straightforward when the last expression is an assignment or control flow.

Assignments return the value being assigned, but 

## Flattening Control Flow
