---
title: "Generating Code for Control Flow Graphs"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating Code for Control Flow Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The purpose of this document is to explain how and why __rstatic__ generates
code when extracting control flow graphs from abstract syntax trees. The
intended audience is compiler developers interested in tweaking the way
__rstatic__ extracts control flow graphs.

<!--
There is also a discussion of how to modify this code generation, with
examples.
-->

## Introduction

For-loops are a high-level control flow structure. In addition to branching to
different program points based on a condition, they automatically set a
variable on each iteration. Thus when the control flow graph is extracted for a
for-loop, some code must be generated to set the loop variable.

Developers may want to customize how code generated for the control flow graph.
For example, there are multiple strategies to generate the code for for-loops;
these are discussed here. More generally, we want to enable developers to
create extensions for the R language and custom languages based on R. These
languages could introduce new control flow structures that require special
handling when the control flow graph is extracted.

Consider a for-loop that adds up the first ten numbers

```{r, eval = FALSE}
x = 0
for (i in 1:10) {
  x = x + i
}
```

To extract the control flow graph, we first need to translate the for-loop into
a while-loop. The while-loop captures the underlying control flow but does not
include any other (implicit) operations. One way we can do this is

```{r, eval = FALSE}
x = 0

iterator = 1:10
counter = 1
while (counter <= length(iterator)) {
  i = iterator[counter]
  counter = counter + 1

  x = x + i
}
```

In this case, we use the variable `iterator` to store the object being iterated
over and `counter` to explicitly count the iterations. Using `iterator` is
important because `1:10` is not a constant, but rather a call to `:`, the
sequence operator. Evaluating `1:10` at each point where we need to use it
would waste CPU cycles. The length of the iterator is explicitly computed in
the condition because in the general case, the value of `iterator` might not be
known until run-time. We could precompute the length and store it in a
variable, but the performance benefit would likely be miniscule.

For for-loops where the iterator is a sequence, such as `1:10` here, there's a
better translation. Instead of computing the entire sequence before the loop
and holding it in memory, we can compute the next element of the sequence at
the beginning of each iteration

```{r, eval = FALSE}
x = 0

i = 1
while (i <= 10) {
  x = x + i

  i = i + 1
}
```

Now the `iterator` and `counter` variables are no longer needed, and the loop
is substanstially simpler. Because the elements are computed one-by-one, this
version of the loop uses less memory than the previous version. Only one
element needs to be held in memory at a time. There's also no time wasted on
computing unnused elements of the iterator if the program exits the loop early.
However, notice that we need explicit information about the beginning, end, and
step size of the sequence at compile time. We do not need to know the exact
values, but if these values are set at run-time, we need to know where to find
them. When the iterator is an arbitrary vector rather than a sequence, we can't
use this strategy at all.

This second strategy also fails if the loop variable `i` is set by the program
within the loop. For instance, consider

```{r, eval = FALSE}
x = 0
for (i in 1:10) {
  i = i^2
  x = x + i
}
```

A literal translation of this code to a while-loop using the second strategy
gives us

```{r, eval = FALSE}
x = 0

i = 1
while (i <= 10) {
  i = i^2
  x = x + i

  i = i + 1
}
```

This is clearly not correct, as `i` will take values `1, 2, 5` and then the
loop will exit due to the condition. Thus we need to use static analysis to
make sure that the loop variable is not set within the loop. In the case where
it is, the `counter` variable is needed to keep track of the current element,
as in

```{r, eval = FALSE}
x = 0

counter = 1
while (i <= 10) {
  i = counter
  counter = counter + 1

  i = i^2
  x = x + i
}
```

This strategy has most of the advantages of the second strategy, but requires
storing two scalar values (the counter and the loop variable) instead of just
one. Like the second strategy, it does not apply when the iterator is not a
sequence.

The functions for extracting control flow graphs don't explicitly translate the
R code for a for-loop into R code for a while-loop. Instead, the translation
happens when the basic blocks for the loop are created. This is because there's
no obvious benefit in constructing the translation as R code only to
immediately extract basic blocks, although it's possible that there are
applications where this would be useful.

One way explicit translation could be used here is to decouple the translation
strategy from control flow graph extraction. However, this would require
additional bookkeeping in order to keep track of the fact that the loop was
originally a for-loop. We want to keep this information may because it
indicates the high-level intent of the program's author, and may be useful in
later stages of analysis and compilation.

Generating basic blocks for a for-loop is similar to generating basic blocks
for a while-loop. First, an entry block, body block, and exit block are created
for the loop. The entry block is where the loop counter and variable are
updated. The entry block terminates with a conditional branch; control flows to
the body block if the loop condition is true, and the exit block if not. The
terminator instruction has the special class `IterTerminator` to indicate that
it's part of a loop rather than an if-statement. Control flow for the body of
the loop is computed recursively. The body block is used as a starting point
for code insertion, but additional blocks are added if there is nested control
flow. Once the control flow graph has been computed for the body of the loop,
the last block is set to branch back to the entry block, which creates a
literal loop (a cycle) in the graph. Finally, the insertion point for the CFG
is set to the exit block, so that extraction can proceed for any code that
comes after the loop.

<!-- FIXME: Give examples of how/why we might modify this procedure -->

<!-- FIXME: Explain how break/next/return are handled -->

<!-- FIXME: Add an example and a diagram -->

<!-- FIXME: Finish this example

## Example: Extending the R Language

Suppose we want to update the loop variable in a loop by calling a closure that
generates the next value. When a closure is used this way, we call it a
_generator_. Generators are useful because they generate elements of the
sequence being iterated over on demand. This avoids the memory cost of storing
the entire sequence and if the loop terminates early, it avoids the cost of
computing the unused elements at the end of the sequence.

We can already use generators in base R with a while-loop, by writing

```{r, eval = FALSE}
i = generator()
while (!is.null(i)) {
  # do something
  # ...
  i = generator()
}
```

However, it would be convenient to simply write

```{r, eval = FALSE}
for (i in generator) {
  # do something
  # ...
}
```

This syntax is preferable because it clearly communicates the programmer's
intention to iterate over the generated values and reduces chance that the
generator will be called at the wrong point in the loop due to programmer
error.

-->

<!-- FIXME: Move to basic discussion of how CFG is extracted

## Control Flow

The order in which expressions in a program are evaluated is called the
_control flow_ of the program. Program control flow is an important piece of
information for most static analysis algorithms. For instance, _constant
propagation_ is a static analysis algorithm that replaces constant variables
(ones whose value does not change over the course of a program) with their
value. This is useful because it improves program performance by eliminating
the run-time cost of looking up the value of the variable. The control flow of
the program determines which values can be propagated.

For example, in the program below, expressions are evaluated in sequence,
starting from the first line. First `42` is assigned to `x`, then the value of
`x` is assigned to `y`. In this program, evaluation proceeds in a straight line
from the first line to the last line. We call this _linear_ control flow. 

```{r, eval = FALSE}
x = 42
y = x
```

Because the first line is always evaluated, the value of `x` that gets assigned
to `y` is always `42`. Thus the constant propagation algorithm can replace `x`
with `42`.

On the other hand, consider the program

```{r, eval = FALSE}
if ()
  x = 42
else
  x = 8

y = x
```

The if-statement in this program

If-statements, and other expressions that only evaluate code when a condition
is met, introduce nonlinear control flow to programs.
-->
