---
title: "rstatic Reference"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rstatic Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette is a reference on abstract syntax trees.

## Introduction

__rstatic__ uses __R6__ classes to represent individual components of R
expressions. The parent class of all components is `ASTNode`, since each
expression is an abstract syntax tree (AST) with the components as nodes.
Abstract syntax trees are structurally similar to the parse trees built into R,
but provide some simplifications to make them more amenable to analysis. 

Note that in contrast to most R objects, __R6__ objects are references. This
means that assigning an __R6__ object to a new variable will not create a copy,
and changes to the new variable will affect the original object. Fields and
methods on an __R6__ object are accessed with the `$` operator.

The `ASTNode` class only has one field and one method. The `parent` field holds
the node's in the abstract syntax tree or control flow graph. The `copy()`
method creates a copy of the node and its children.

## Literals

Literal numbers, strings, logicals and nulls are represented by subclasses of
`Literal`.

In the example below, the literal `6.28` is converted to a code object. The
`quote_ast()` function is like `to_ast()`, but automatically quotes its argument.
```{r}
tau = quote_ast(6.28)

tau
```
In this case, the result is a `Numeric` object, which is a subclass of
`Literal`.
```{r}
class(tau)
```
The `value` field stores the `Literal`'s value.
```{r}
tau$value
```

## Symbols & Parameters

Variable names or _symbols_ are represented by the `Symbol` class.

The code below creates a new symbol named `x`.
```{r}
x = Symbol$new("x")

x
```

`Symbol`s have fields `basename`, `ssa_number`, and `name`. The `basename`
field is the name of the symbol as it appears in the original R program.
```{r}
x$basename
```

The `ssa_number` field is the SSA number assigned to this symbol or `NA` if
none was assigned.
```{r}
x$ssa_number
```

The `name` field is read-only and contains the SSA name of the symbol. When
`ssa_number` is `NA`, `name` is the same as the `basename`.
```{r}
x$name
```
```{r}
x$ssa_number = 3

x$name
```

Function parameters are represented by the `Parameter` class, a subclass of
`Symbol`. In addition to inherited members, `Parameter`s have a `default` field
for storing the default argument to the parameter.

As an example, the code below creates a parameter called `bins` with default
argument `8`.
```{r}
Parameter$new("bins", Integer$new(8))
```
Note that the R representation of a lone parameter is a pairlist with one
element, which is why the parameter is displayed as a pairlist.

## Assignments

Assignments with `=` or `<-` are represented by the `Assign` class, which has
fields `read` and `write`. The `write` field holds the left-hand side of the
assignment, and the `read` field holds the right-hand side.

As an example, consider the code below, which assigns the complex number `3+2i`
to the variable `z`.
```{r}
assign = quote_ast(z <- 3+2i)
```

Here, the `write` field holds a symbol and the `read` field holds a complex
value.
```{r}
assign$write
```
```{r}
assign$read
```

The setter methods should be used to set `write` and `read`, rather than
setting the fields directly. They ensure that children of the `Assign` node
have the `parent` field set appropriately.

## Function Calls

Expressions with arguments are represented by the `Application` class. Calls
and return statements are represented by `Call` and `Return`, which are
subclasses of `Application`.

For instance, the snippet below creates a call to `sum()` with no arguments.
```{r}
call_sum = Call$new("sum")
```

All `Application`s have an `args` field that holds the list of arguments. For
`call_sum`, this is an empty list.
```{r}
call_sum$args
```
We can set the sum in the example to have two arguments, the integers `4` and
`6`.
```{r}
call_sum$args = list(Integer$new(4), Integer$new(6))

call_sum
```

In addition to the members inherited from `Application`, `Call`s also have a
`fn` field that holds the function being called, either as a `Symbol` or as a
`Function` (for calls to anonymous functions).

In the sum example, the `fn` field is a symbol named `sum`.
```{r}
call_sum$fn
```

Specialized calls such as replacement functions and `.Internal()` are
represented by subclasses of `Call`. This makes it easy to use S3 dispatch to
treat them like any other `Call` or to handle them as special cases.

The `Replacement` class represents a call to a replacement function. This class
does not add any new members, but requires that the function name `fn`
ends with `"<-"`.

The `Internal` class, which represents a call to `.Internal()`, does not add
any new members or functionality. It exists only to distinguish these calls
from others.

## Braces

Parentheses and curly braces are represented by the `Brace` class. This class
has fields `body` and `is_paren`. The `is_paren` field is `TRUE` if the `Brace`
object represents parentheses and `FALSE` if the object represents curly
braces.

As an example, the code shown below converts a parenthetical expression to a
`Brace`.
```{r}
paren = quote_ast( (2 + 2) )

paren
```

The `body` field holds the list of expressions inside the `Brace`. For
parentheses, this list must have length 0 or 1 to be valid. The list can be any
length for curly braces.
```{r}
paren$body
```

## Functions & Primitives

Callable objects are represented by the `Callable` class. The `Callable` class
defines a `params` field that holds a list of parameters (and default
arguments) for the callable object.

`Function` and `Primitive`, both subclasses of `Callable`, represent functions
and primitives, respectively. The `Function` class adds a `body` field to the
members inherited from `Callable`. The `body` field holds the single expression
inside the function; for functions that contain more than one expresion, `body`
holds a `Brace` object.

For example, the code below converts a function to a `Function` object.
```{r}
to_celsius = function(x) {
  (x - 32) * 5 / 9
}

to_celsius = to_ast(to_celsius)

to_celsius
```

The parameter `x` is in the `params` field.
```{r}
to_celsius$params
```

The code inside the function definition is in the `body` field.
```{r}
to_celsius$body
```

Since primitives have names but do not contain any R code, the `Primitive`
class adds a `fn` field to the members inherited from `Callable`. The `fn`
field holds the name of the primitive, and is similar to the `fn` field on the
`Call` class.

## Control Structures

Control structures are represented by the `Next`, `Break`, `If`, `For`, and
`While` classes. `Next` and `Break` do not add any new functionality to
`ASTNode`, and are merely meant for method method dispatch.

The `If` class has fields `condition`, `true`, and `false`. These fields hold
to the if-statement's condition, "if" branch, and "else" branch, respectively.
The `false` field is `NULL` for if-statements that don't have an "else" branch.

For instance, the code below shows the AST for an if-statement.
```{r}
ast = quote_ast(
  if (x > 0) {
    x = x + sqrt(x)
  })

ast
```

The condition on the if-statement is `x > 0`, and the `condition` field
reflects this.
```{r}
ast$condition
```

The body of the if-statement is held in the `true` field. The `false` field is
`NULL` since the if-statement does not have an "else" branch.
```{r}
ast$true
```

The `For` class has fields `ivar`, `iter`, and `body`. The `ivar` field holds
the iteration variable, the `iter` field holds the object being iterated over,
and the `body` field holds the list of expressions that make up the body of the
loop.

The `While` class has fields `condition`, `body`, and `is_repeat`. Similar to
the `If` class, the `condition` field holds the condition for the loop. The
`body` field holds the list of expressions that make up the body of the loop.
The `is_repeat` field is `TRUE` if the loop is a `repeat` loop and `FALSE` if
the loop is a `while` loop. When `is_repeat` is `TRUE`, the `condition` field
is ignored (and defaults to `NULL`).
