---
title: "Introduction to rstatic"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rstatic Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette introduces the __rstatic__ API for accessing information about an
program.

## Getting Started With Code Analysis

The __rstatic__ package is all about extracting information from R code. We can
use this information in various ways: to check for programming errors, to
estimate hardware requirements, to gain a better understanding of how the code
is organized (especially for unfamiliar code), to generate additional code, or
to compile.

Code is typically written and stored as a string of text. Strings don't expose
any information about structures within the code, so the first step in analysis
is to _parse_ the code into an _intermediate representation_. Since the purpose
of an intermediate representation is to make code easier to explore and
manipulate, many different intermediate representations exist. Choosing the
right intermediate representation for a given task can make the task much
easier, so many code tools, including this package, support more than one
intermediate representation.

The __rstatic__ package can convert R code to two different intermediate
representations: abstract syntax trees and control flow graphs. In an _abstract
syntax tree_, each node in the tree corresponds to an expression. When an
expression is composed of subexpressions, these appear as its children in the
tree.

This is similar to a _parse tree_, the R interpreter's built-in representation
for code. The main difference is that abstract syntax trees abstract away some
of the syntactic details that get retained in a parse tree. Syntax is usually
not as important for analysis as the meaning, or semantics, of the code. In an
R parse tree, the class of each node is determined by the syntax of the
expression the node corresponds to. For example, any expression with call
syntax `f()` is represented by a node with class `call`. This includes the
`return()` statement, even though the meaning of the return statement is quite
different from an ordinary function call.


## Example 1: Interacting With Code

Let's start by examining the AST for an R implementation of the quadratic
formula. The `toAST()` function converts functions and quoted R expressions to
abstract syntax trees.
```{r}
quad = function(a, b, c) {
  rad = sqrt(b^2 - 4*a*c)
  roots = (-b + c(rad, -rad)) / 2*a

  return (roots)
}

library(rstatic)

ast = toAST(quad)

ast
```

Printing `ast` reveals that it's a `Function` object and shows the code inside.
For convenience, descendants of `ASTNode` list all field and method names when
printed.
```{r}
class(ast)
```

The contents of the function are in the `body` field. Since the function has
multiple lines, they're wrapped in curly braces. Thus the `body` field holds a
`Brace` object.
```{r}
braces = ast$body

braces
```

`Brace` objects also have a `body` field, which lists the lines being wrapped.
For this example, there are three lines: two assignments and a return
statement, represented by `Assign` and `Return` objects, respectively.
```{r}
braces$body
```

Inspecting the first `Assign`, we see that it has members `read` and `write` in
addition to the inherited `ASTNode` members. The `write` field holds the
variable being assigned (that is, the left side of the assignment).
```{r}
lhs = braces$body[[1]]$write

lhs
```

Drilling down further, `lhs` holds a `Symbol`, which has fields `basename`,
`ssa_number`, and `name` in addition to the inherited members. The `basename`
field is original name of the variable in the R program, as a string.
```{r}
lhs$basename
```

The `ssa_number` and `name` fields are used for SSA form, which is discussed in
a subsequent section of this vignette. We can change `basename` to change the
name of the variable in the code.
```{r}
lhs$basename = "radical"

ast
```


<!-- SEE THE R/ASTTraverse.R code which is a more convenient way to do this. 
Good to explain you can do this also, but best to lead with the astTraverse() function
and some examples.  
-->
ASTs are cumbersome to manipulate by hand. In practice, the easiest way to
apply operations to a whole AST is to write a recursive S3 method. For
instance, if we wanted to rename all instances of a variable, we could use the
method below:
```{r}
replace_var = function(node, old, new) {
  UseMethod("replace_var")
}

replace_var.Symbol = function(node, old, new) {
  if (node$basename == old)
    node$basename = new
  invisible(NULL)
}

replace_var.Brace =
function(node, old, new) {
  lapply(node$body, replace_var, old, new)
  invisible(NULL)
}

replace_var.Application = function(node, old, new) {
  lapply(node$args, replace_var, old, new)
  invisible(NULL)
}

replace_var.Function = function(node, old, new) {
  replace_var(node$body, old, new)
  invisible(NULL)
}

replace_var.Assign = function(node, old, new) {
  replace_var(node$read, old, new)
  replace_var(node$write, old, new)
  invisible(NULL)
}

replace_var.Literal = function(node, old, new) { }
```

This makes it easy to change `rad` to `radical` globally:
```{r}
replace_var(ast, "rad", "radical")

ast
```

There are a few things worth noting in this example. First, the AST is updated
in place rather than returned by `replace_var`. The `NULL` return values are
not strictly necessary, but emphasize that `replace_var` is used only for its
side effects. Although it wasn't necessary here, it's often useful to define a
helper __R6__ class to hold additional state during recursive tree traversal.

Second, the method had to be defined for every class of node that could appear
in the tree. This boilerplate is a tradeoff for flexibility. Intermediate
classes such as `Application` (for expressions that have arguments) and
`Literal` (for literals) help to reduce the boilerplate. For convenience, the
package may have more intermediate classes in the future.

Third, in this example there's no real benefit, other than clarity, in
traversing the AST rather than the parse tree for the code. This is usually the
case for operations that don't use information about each node's context in the
tree. On the other hand, for operations that do use contextual information, the
AST makes it possible to inspect and even modify ancestors the current node.

Finally, the method could be vectorized in `old` and `new` by changing
`replace_var.Symbol()`.

## Control Flow Graphs

The benefits of working with ASTs over parse trees are relatively small. Most
of the analysis passes in __rstatic__ operate on control flow graphs, which are
convenient because they make control flow explicit and use a common language
(graph edges) for all control flow structures.

The function shown below simulates a two-state Markov chain for `n` steps. The
control flow is relatively simple, with just one for-loop and one return
statement.
```{r}
markov2 = function(n, p, init = 0) {
  x = c(init, numeric(n))
  for (i in 1:n) {
    idx = x[i] + 1
    x[i + 1] = rbinom(1, 1, p[idx])
  }

  return (x[-1])
}
```

We can convert the `markov2` function to a control flow graph with the
`toCFG()` method. When given a function as input, `toCFG()` will
automatically convert it to an AST and then convert the AST to a CFG.
```{r}
markov2 = toCFG(markov2)

cfg = markov2$cfg
class(cfg)
```

As you might expect, the `ControlFlowGraph` class represents a control flow
graph. In __rstatic__, each control flow graph corresponds to one function;
`toCFG()` automatically wraps code in a function definition as needed.
`ControlFlowGraph` is an __R6__ class, so the `cfg` variable is a reference, as
discussed in the previous section.

`ControlFlowGraph` is a subclass of `FlowGraph`, which represents flow graphs
in general. The `FlowGraph` class has several fields, but the most important
ones are `blocks` and `graph`. The `blocks` field holds a list with one
element, or _block_, for each vertex in the graph. The names on the list match
the vertex names on the graph. Each block holds metadata about the
corresponding node. For convenience, elements of the `blocks` field can be
accessed by name or index with the extraction operator `[[` on the `FlowGraph`
object.

For instance, the code below gets the first block in `cfg`.
```{r}
cfg[[1]]
```

The `graph` field holds the actual graph, as an __igraph__ graph. Calling
`plot()` on a `FlowGraph` object displays a visualization of the graph.
```{r}
cfg$graph

plot(cfg)
```

In a control flow graph, the vertices are blocks of code that execute linearly,
without any branches in control flow. These vertices are sometimes called
_basic blocks_. At the end of each basic block, control flows to another basic
block, which may depend on a condition. The directed edges in the control flow
graph show how control flows from block to block.

The blocks in the `ControlFlowGraph` class are basic blocks, which are
represented by the `BasicBlock` class. The `BasicBlock` class has fields
`body`, `name`, `phi`, and `terminator`. The `name` field holds the block's
name, which is the same as the vertex name. By default, __rstatic__ uses a "%"
followed by an integer for block names. The "%" helps to avoid confusion with
integer indices.
```{r}
cfg[[1]]$name
```

The `body` field holds a list of `ASTNode` objects, one for each ordinary
expression in the block.
```{r}
cfg[[1]]$body
```

The `terminator` field holds the final expression in the block, which is always
a branch to a different block or a return from the function the CFG represents.
These are represented by subclasses of the `Terminator` class, which are
discussed in more detail in a subsequent section.
```{r}
cfg[[1]]$terminator
```

The `phi` field holds a list of `Phi` objects. These are related to SSA and
appear at the beginning of the basic block when the block is printed. The `Phi`
class is discussed in the "Static Single-Assignment Form" section.

### Terminators

The `Terminator` class has only one method, `successors()`, which returns the
indices of the basic blocks that control flows to from the terminator.

<!--
### Loops


## Static Single-Assignment Form


## Basic Analysis Functions
-->


<!--

The following example simulates a betting strategy where the player doubles
their next bet every time they win and halves their next bet every time they
lose. Suppose we wanted to 

```{r, eval = FALSE}
n = 100
winnings = 64
bet = 2

for (i in 1:n) {
  if (bet > winnings)
    break

  is_win = rbinom(1, 1, 0.5)
  if (is_win == 1) {
    winnings = winnings + bet
    bet = 2 * bet
  } else {
    winnings = winnings - bet
    bet = bet / 2
  }
}
```

The parse tree doesn't 
-->

